---
title: 'Task 1: PCA (PELEGRI)'
author: "Patrick Pelegri-O'Day"
date: "2/13/2022"
output:
  html_document: 
    theme: flatly
    highlight: pygments
    code_folding: hide
---

### Overview
This brief describes the relationship among environmental variables across all countries of the world using a PCA analysis. The data are compiled from publicly available remote sensing datasets. Most of the data is derived by calculating the mean for each country at a reduction scale of about 10km.

**Citation:** Compiled and provided by @zander_venter on Kaggle. The data are acquired from Google Earth Engine (https://earthengine.google.com/).

### Setup
```{r setup, include=TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(ggfortify)
library(patchwork)
```

**Variables selected for PCA analysis**

- Accessibility to cities
- Tree canopy cover
- Accessibility to cities
- Tree canopy cover
- Rainfall driest month
- Rainfall wettest month
- Mean annual rainfall
- Rain seasonality
- Temperature annual range
- Temperature diurnal range
- Mean annual temperature
- Temperature seasonality

**NA values:** 12 countries are dropped due to NA values, primarily due to countries that do not have data on tree canopy cover.

Read in data
```{r read in data}
env_df <- read_csv(here('data', 'world_env_vars.csv'))
```

Select variables of interest and drop NAs
```{r}
env_sub_df <- env_df %>% 
  select(Country, accessibility_to_cities, tree_canopy_cover, rain_driestMonth,rain_wettestMonth, rain_mean_annual, rain_seasonailty, temp_annual_range, temp_diurnal_range, temp_mean_annual, temp_seasonality) %>% 
  drop_na() %>%  # Variable with the most NAs is wind with 11. The others have 5 or 6
  rename( # for display in plot
    "Accessibility to cities" = accessibility_to_cities,
    "Tree canopy cover" = tree_canopy_cover,
    "Rainfall driest month" = rain_driestMonth,
    "Rainfall wettest month" = rain_wettestMonth,
    "Mean annual rainfall" = rain_mean_annual,
    "Rain seasonality" = rain_seasonailty,
    "Temperature annual range" = temp_annual_range,
    "Temperature diurnal range" = temp_diurnal_range,
    "Mean annual temperature" = temp_mean_annual,
    "Temperature seasonality" = temp_seasonality
  )
```


```{r create pca object}
env_pca <- env_sub_df %>% 
  select(-Country) %>% # drop country because all variables for pca need to be numeric
  prcomp(scale = TRUE) # scale the variables so they're directly comparable

```

### Analysis

Create biplot
```{r}
autoplot(env_pca,
         data = env_sub_df,
         x = 1,
         y = 2,
         colour = "lightyellow3",
         loadings = TRUE,
         loadings.label = TRUE, # show labels of original vectors
         loadings.colour = 'slategray3',
         loadings.label.colour = 'firebrick4',
         loadings.label.vjust = -0.4,
         loadings.label.size = 3) +
  theme_minimal()
```
<br>
**PCA plot 1 takeaways:**

- Tree canopy was positively correlated mean and wettest month rainfall, while negatively correlated with temperature seasonality and temperature range. This aligns with our knowledge that trees need a high level of precipitation to thrive. While we cannot infer causality from the correlation, it raises an interesting questions about if the presence of forests serves as a buffer against temperature extremes or if temperature extremes undermine trees' survivability.
- Driest month of rainfall was negatively correlated with temperature range, perhaps pointing to the fact that desert-like environments have large temperature swings throughout the day.
- Rain seasonality was almost completely unrelated to temperature seasonality.
- The variation in accessibility to cities was not captured very well by PCA 1 or PCA 2, indicating a weak relationship between accessibility to cities and the other variables analyzed.

Create screeplot
```{r}
sd_vec <- env_pca$sdev
var_vec <- sd_vec^2

pc_names <- colnames(env_pca$rotation)
pct_expl_df <- data.frame(v = var_vec,
                          pct_v = var_vec / sum(var_vec),
                          pc = fct_inorder(pc_names)) %>% 
  mutate(pct_lbl = paste0(round(pct_v * 100, 1), '%'))

ggplot(data = pct_expl_df, aes(x = pc, y = v)) +
  geom_col(width = 0.8, fill = 'cornsilk4') + 
  geom_text(aes(label = pct_lbl, vjust = -0.25, nudge_y = 0.0005)) +
  labs(x = "\nPrincipal component", y = "Variance estimated\n") +
  theme_minimal()
```
<br>
**Screeplot takeaways:**

 - The first two PCAs capture ~70% of the variance.
 - The most visible "elbows" in the screeplot occur after the third PCA or fourth PCA lis accounted for. In order to show the PCAs that occur before the "elbow" and a set of PCAs that add up to over 80%, an additional PCA plot is shown below comparing PCA 1 to PCA 3.
 
Create second biplot
```{r}
autoplot(env_pca,
         data = env_sub_df,
         x = 1,
         y = 3,
         colour = "lightyellow3",
         loadings = TRUE,
         loadings.label = TRUE,
         loadings.colour = 'slategray3',
         loadings.label.colour = 'firebrick4',
         loadings.label.vjust = -0.6,
         loadings.label.size = 3) +
  theme_minimal()
```
<br>
**PCA plot 2 takeaways:**

- The variance of accessibility to cities is better represented by PCA 3.
- Accessibility of cities is quite uncorrelated with the other variables.
- The strongest correlation with accessibility to cities is mean annual temperature.
