---
title: 'Task 2: Hierarchical clustering (PELEGRI)'
author: "Patrick Pelegri-O'Day"
date: "2/14/2022"
output:
  html_document: 
    theme: flatly
    highlight: pygments
    code_folding: hide
---

### Overview

This report uses hierarchical clustering by complete linkage to create a dendogram showing multivariate clustering for water chemistry by site. A tanglegram is generated to show how clustering using single linkage and complete linkage differ. The data are drawn from six Santa Barbara Coastal Long Term Ecological Research (LTER) sites. Monitoring has been ongoing since year 2000.

**Citation:** Santa Barbara Coastal LTER and J. Melack. 2019. SBC LTER: Land: Stream chemistry in the Santa Barbara Coastal drainage area, ongoing since 2000 ver 16. Environmental Data Initiative. https://doi.org/10.6073/pasta/67a558a24ceed9a0a5bf5e46ab841174. 

### Setup

```{r setup, include=TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(here)
library(tidyverse)
library(naniar)
library(dendextend)
library(ggdendro)
```

```{r read in data}
water_raw_df <- read_csv(here('data', 'sbc_lter_registered_stream_chemistry.csv'))
```

Prepare data for clustering
```{r drop NAs}
# Drop NAs
water_dropna_df <- water_raw_df %>% 
  replace_with_na_all(condition = ~.x == -999.0) %>% # replace values of -999.0 with 'NA' in all columns
  drop_na()
```

To perform hierarchical clustering by site, we need to reformat the data so that the data are arranged in a single summary row per site based on means from all observations at that site.
```{r summarize and scale data}
# Group by site code and summarize to get mean of each variable
water_summarize_df <- water_dropna_df %>% 
  group_by(site_code) %>% # for each site...
  summarize_all(~mean(.x))  # ...take mean of each column
  
# Remove character variables and scale numerical values  
water_scaled_df <- water_summarize_df %>% 
  select(-timestamp_local, -site_code) %>% # drop character variables
  scale() # make column vectors comparable in scale

# Make site code row name
rownames(water_scaled_df) <- water_summarize_df$site_code 
```

Calculate distance matrix using Euclidean distance
```{r}
euc_distance <- dist(water_scaled_df, method = 'euclidean', diag = TRUE, upper = TRUE)
```

### Clustering analysis

Complete linkage agglomerative clustering
```{r}
hc_complete <- hclust(euc_distance, method = 'complete')

# Don't include plot in final
# plot(hc_complete, cex = 0.6, hang = -1)
```

Single linkage agglomerative clustering
```{r}
hc_single <- hclust(euc_distance, method = 'single')

# Don't include plot in final
# plot(hc_single, cex = 0.6, hang = -1)
```


```{r}
ggdendrogram(hc_complete, rotate = TRUE) +
  theme_minimal() +
  labs(x = 'Compound\n', y = 'Distance')
```
<br>
**Figure 1:** Dendrogram of agglomerative clustering by complete linkage for water chemistry by site in Santa Barbara Coastal LTER based on data from 2000 to 2005.

<br>
Make tanglegram to compare clustering from single and complete linkage
```{r}
### convert to class dendrogram
dend_complete <- as.dendrogram(hc_complete)
dend_single <- as.dendrogram(hc_single)

tanglegram(dend_complete, dend_single)
```


